\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-05 paper/ms.tex
Completed prediction section.

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
of the covariance matrix:
\begin{equation}
\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
the likelihood with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
&=& \bvec{q}^T \bvec{q}.
\end{eqnarray}
The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.

@@

Since the covariance matrix may be written as the sum of exponentials, we can write:
\begin{eqnarray}
\bar f_{*,i} &=& K(t_{*,i},\bvec{t}) \bvec{z} \\
&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t_{*,i}-t_n\vert}\left(a_j \cos{(d_j \vert t_{*,i}-t_n\vert)} +b_j \sin{(d_j \vert t_{*,i}-t_n\vert)}\right) z_j.
\end{eqnarray}
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,i}$ and
$\{t_{n0+1},...,t_N\} \ge t_{*,i}$, gives
\begin{eqnarray}
\bar f_{*,i} &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t_{*,i}-t_n)} \left[a_j \cos{(d_j(t_{*,i}-t_n))}+b_j\sin{(d_j(t_{*,i}-t_n))}\right] z_n \\
&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t_{*,i})} \left[a_j \cos{(d_j(t_n-t_{*,i}))}+b_j\sin{(d_j(t_n-t_{*,i}))}\right] z_n.
\end{eqnarray}
%Each of these double-sums may have the sums interchanged, giving:
%\begin{eqnarray}
%\bar f_{*,i} &=& Re\left[\sum_j e^{-(c_j+id_j)( t_{*,i}-t_{*,i-1})} \sum_{n=1}^{n_0} (a_j+ib_j) e^{-(c_j+id_j)( t_{*,i-1}-t_n) } z_n\right. \\
%&+&\left.
%\sum_j e^{-(c_j+id_j)( t_{*,i+1}-t_{*,i})} \sum_{n=n_0+1}^{N} (a_j+ib_j) e^{-(c_j+id_j)( t_n-t_{*,i+1}) } z_n\right]
%\end{eqnarray}
We can compute this recursively.  Defining
\begin{eqnarray}
Q^-_{i,n_0,2j-1} &=& \sum_{n=1}^{n_0} a_j e^{-c_j(t_{*,i}-t_n)} \cos{(d_j(t_{*,i}-t_n))} z_n,\\
Q^-_{i,n_0,2j} &=& \sum_{n=1}^{n_0} b_j e^{-c_j(t_{*,i}-t_n)} \sin{(d_j(t_{*,i}-t_n))} z_n,\\
Q^+_{i,n_0,2j-1} &=& \sum_{n=n_0+1}^N a_j e^{-c_j(t_n-t_{*,i})}\cos{(d_j(t_n-t_{*,i}))} z_n,\\
Q^+_{i,n_0,2j} &=& \sum_{n=n_0+1}^N b_j e^{-c_j(t_n-t_{*,i})} \sin{(d_j(t_n-t_{*,i}))} z_n,
\end{eqnarray} gives:
\agoltodo{Convert the following to real, \celerite\ notation .}
\begin{equation}
\bar f_{*,i} = \sum_p e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i-1,n_0,p} + \sum_p e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i+1,n_0,p}.
\end{equation}
Once $Q^-_{1,p}$ and $Q^+_{N,p}$ are obtained in $\mathcal{O}(JN)$ operations, then the subsequent $Q^-_{i,p}$ and $Q^+_{i,p}$ values can be obtained in $\approx \mathcal{O}(JM)$ operations
\begin{eqnarray}
Q^-_{i,p} &=& e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i,p},\\
Q^+_{i,p} &=& e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i,p},
\end{eqnarray}
for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.

The variance evaluated at future predicted times $\bvec{t_*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
\begin{eqnarray}
\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t_{*,i+1})\\
&=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
&\times& e^{-\beta_p(t_{*,i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}\\
&=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
\end{eqnarray}
for $1 \le p,q \le J$,
where
\begin{equation}
P_{i,p,q} = e^{-\beta_p(t_{*,i}-\bvec{t}^T)}
K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}.
\end{equation}
For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
with the Cholesky decomposition.
Subsequent $P_{i+1,p,q}$ may be obtained recursively as
\begin{equation}
P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t_{*,i+1}- t_{*,i})} P_{i,p,q},
\end{equation}
in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
Note that this variance formula {\it only} works for prediction; we have
yet to derive a formula for which $\bvec{t_*}$ and $\bvec{t}$ are
interspersed.

>>>>

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
of the covariance matrix:
\begin{equation}
\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that $\mathcal{Q} = -2 \mathcal{L}$
with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
&=& \bvec{q}^T \bvec{q}.
\end{eqnarray}
The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$.

@@

Since the covariance matrix may be written as the sum of exponentials, the predicted mean is given by
\begin{eqnarray}
\bar f_{*,m} &=& K(t_{*,m},\bvec{t}) \bvec{z} \\
&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t_{*,m}-t_n\vert}\left(a_j \cos{(d_j \vert t_{*,m}-t_n\vert)} +b_j \sin{(d_j \vert t_{*,m}-t_n\vert)}\right) z_j,
\end{eqnarray}
for $m =1$ to $M$.
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,m}$ and
$\{t_{n0+1},...,t_N\} \ge t_{*,m}$, gives
\begin{eqnarray}
\bar f_{*,m} &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t_{*,m}-t_n)} \left[a_j \cos{(d_j(t_{*,m}-t_n))}+b_j\sin{(d_j(t_{*,m}-t_n))}\right] z_n \\
&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t_{*,m})} \left[a_j \cos{(d_j(t_n-t_{*,m}))}+b_j\sin{(d_j(t_n-t_{*,m}))}\right] z_n.
\end{eqnarray}
We can compute this recursively.  Defining
\begin{eqnarray}
Q^-_{n_0,i} &=& \left(Q^-_{n_0-1,i}+z_n\tilde V_{n_0,i}\right)e^{-c_j(t_{n+1}-t_n)},\\
X^-_{m,n_0,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i},\\
Q^+_{n_0,i} &=& \left(Q^+_{n_0+1,i}+z_n\tilde U_{n,i}\right)e^{-c_j(t_n-t_{n-1})},\\
X^+_{m,n,i} &=& e^{-c_j(t_{n-1}-t_m^*)} \tilde V_{m,i},
\end{eqnarray}
where $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$ to $2J$, and in $X^\pm_{m,n,i}$ the expressions
for $\tilde U_{m,i}$ and $\tilde V_{m,i}$ are evaluted at $t_{m}^*$.
Two sweeps must be carried out, upward in $n$ and $m$ to compute $Q^-$ recursively, and downward in
$m$ and $n$ to compute $Q^+$ recursively, for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.
For each value of $m$, $Q^\pm$ are recursively updated over $n$ until $n_0$ is reached, at which
point the predicted mean can be computed from
\begin{equation}
\bar f_{*,i} = \sum_{i=1}^{2J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0,i}X^+_{m,n_0,i}\right],
\end{equation}

%The variance evaluated at future predicted times $\bvec{t_*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
%\begin{eqnarray}
%\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t_{*,i+1})\\
%&=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
%&\times& e^{-\beta_p(t_{*,i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}\\
%&=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
%\end{eqnarray}
%for $1 \le p,q \le J$,
%where
%\begin{equation}
%P_{i,p,q} = e^{-\beta_p(t_{*,i}-\bvec{t}^T)}
%K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}.
%\end{equation}
%For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
%with the Cholesky decomposition.
%Subsequent $P_{i+1,p,q}$ may be obtained recursively as
%\begin{equation}
%P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t_{*,i+1}- t_{*,i})} P_{i,p,q},
%\end{equation}
%in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
%Note that this variance formula {\it only} works for prediction; we have
%yet to derive a formula for which $\bvec{t_*}$ and $\bvec{t}$ are
%interspersed.
